\large End-to-End Verification of Initial and Transition Properties of GR(1) Designs in SPARK

### 2020-05-28 doc/2020-SEFM/paper.tex
Incorporating most recent paper comments

% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Institute 1 \\
\email{author1@email.com} \and
Institute 2 \\
\email{author2@email.com}  \and
Institute 3 \\
 \email{author3@email.com}}
%
%\institute{Air Force Research Laboratory, WPAFB, OH 45433, USA \\
%\email{laura.humphrey@us.af.mil} \and
%LinQuest Corp., Beavercreek, OH 45431, USA \\
%\email{james.hamil.ctr@us.af.mil}  \and
%AdaCore, F-75009 Paris, France \\
% \email{huguet@adacore.com}}
%
\maketitle              % typeset the header of the contribution
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 150--250 words.
Manually designing control logic for reactive systems is time-consuming and error-prone. 
An alternative is to automatically generate controllers using ``correct-by-construction'' synthesis approaches. 
Recently, there has been interest in synthesis from Generalized Reactivity(1) or GR(1) specifications, 
since the required computational complexity is relatively low, and several tools now exist for GR(1) synthesis.
However, while these tools implement synthesis approaches that are theoretically ``correct-by-construction,'' 
errors in tool implementation can still lead to errors in synthesized controllers. 
We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications.
Toward this end we have modified Salty -- a tool that produces executable software implementations of controllers 
from GR(1) specifications in a variety of programming languages -- to produce implementations in SPARK.
SPARK is both a programming language and associated set of verification tools, so it has the potential to enable the ``end-to-end'' verification we desire.
In this paper, we discuss our experience to date using SPARK to implement controllers and verify them against 
a subset of properties comprising GR(1) specifications, namely system initial and transition properties. 
We also discuss lessons learned about how to best encode controllers in SPARK for verification, 
examples in which verification found unexpected controller behaviors, and 
caveats related to the interpretation of GR(1) specifications.

@@

Reactive systems must be capable of correctly responding to various inputs, 
e.g. originating from human users or events in the system's operational environment. 
The process of manually designing the control logic for such systems is both time-consuming and error prone. 
An alternative is to use ``correct-by-construction'' synthesis approaches to automatically generate 
the design of a system's control logic directly from specifications, which can 
reduce both design time and the likelihood of errors 
\cite{alur2016compositional}, \cite{fainekos2009temporal}, \cite{guo2014cooperative}, \cite{kupermann2001synthesizing}. 
In general, \emph{synthesis} is the process of automatically generating a design from a specification. 
More specifically, \emph{reactive synthesis} approaches generate designs in the context of an uncontrolled environment, 
assumptions about which are encoded in the specification.
Reactive synthesis approaches tend to have high computational complexity,  
so there is particular interest in synthesis from Generalized Reactivity(1) or GR(1) specifications, 
the complexity of which is only polynomial in the size of the game graph encoded by the specification \cite{bloem2012}.
Synthesis from GR(1) specifications has been used to generate digital circuits \cite{ehlers2012symbolically} 
and controllers for teams of unmanned vehicles \cite{apker2016}, ground robots \cite{kress2007s}, 
software-defined networks \cite{wang2013automated}, and aircraft power distribution systems \cite{xu2012case}, to name a few.

A GR(1) specification $\varphi$ takes the form $\varphi = \varphi^e \limplies \varphi^s$, 
where $\varphi^e$ encodes assumptions about the \emph{environment} in which a system is to operate,
and $\varphi^s$ encodes guarantees the \emph{system} should make under those assumptions \cite{Ehlers2016}. 
More specifically, $\varphi$ takes the form 
$\varphi = (\varphi^e_i \land \varphi^e_t \land \varphi^e_l)  \limplies (\varphi^s_i \land \varphi^s_t \land \varphi^s_l)$,
where $\varphi^e_i$ and $\varphi^s_i$ are \emph{initial} properties, 
$\varphi^e_t$ and $\varphi^s_t$ are \emph{transition} or safety properties, and 
$\varphi^e_l$ and $\varphi^s_l$ are \emph{liveness} properties. 
For inputs in the set $\inputs$ controlled by the environment 
and outputs in the set $\outputs$ produced by the system, terms are:

@@

Several tools for GR(1) synthesis are available. 
For example, RATSY %(Requirements Analysis Tool with Synthesis) 
\cite{bloem2010ratsy} has a focus on circuit design and synthesizes designs 
encoded in BLIF, %(Berkeley Logic Interchange Format),
Verilog, and HIF. %(HDL Intermediate Format). 
Similarly, Anzu \cite{jobstmann2007anzu} produces circuit designs in Verilog.
LTLMoP 
%(Linear Temporal Logic MissiOn Planning)  
\cite{finucane2010ltlmop} is focused on control of robots modeled as hybrid systems, 
and it synthesizes designs as hybrid controllers with handler modules to help connect controllers to simulated or real-world systems. 
TuLiP 
%(Temporal Logic Planning) 
\cite{TuLiP2011} has a similar focus and can synthesize controller implementations in Python. 
Slugs 
%(SmalL bUt Complete GROne Synthesizer) 
\cite{Ehlers2016} is architected to allow users to tailor synthesis algorithms,
 e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and it produces mathematical representations of controller designs.
 %% -- TODO: put more details about input types that Salty allows? About what optimizations it does?
Salty \cite{elliott2019salty} provides a front-end to Slugs that makes specifications easier to write and debug and 
a back-end that turns controller designs into executable software implementations in a variety of programming languages, 
including Python, C++, Java, and now SPARK. 
Though these tools implement synthesis algorithms that are theoretically ``correct-by-construction,''
tool implementation errors could still result in errors in synthesized controllers.
We are therefore interested in ``end-to-end'' verification of synthesized controllers.
Toward this end, we have extended Salty to produce software implementations of synthesized controller designs in SPARK.

SPARK, which is based on the Ada programming language, is both a programming language 
with a specification language and an associated verification toolset \cite{hoang2015spark}.
SPARK translates code to the Why language for processing by the Why3 theorem prover, 
and Why3 uses other automated provers -- including SMT solvers Z3, CVC4, and Alt-Ergo -- to discharge verification conditions.
Though SPARK aims to perform fully automated verification, manual adaptation of the source code is in general necessary. 
For example, if one were to translate source code originally written in Ada to SPARK, 
one would have to remove unsupported features such as functions with side-effects, aliased variable names, exception handlers, etc.  
Formal verification with SPARK also requires supplying requirements at the level of the source code in the 
form of contracts, e.g. preconditions and postconditions that summarize the assumptions and guarantees provided by indvidual subprograms.
There are different levels to which one can use SPARK to verify a program. 
Generally, the level of verification performed during development is incremental, going from lowest to highest. 
Verification at the lowest level, ``stone level,'' is achieved when code is accepted by SPARK, 
since the SPARK language has stricter legality rules than Ada. 
Verification at the next level, ``bronze level,'' is achieved when flow analysis returns with no error; 
flow analysis checks for errors such as reads of unintialized data or violations of user-specified data flow contracts. 
Verification at the next level, ``silver level,'' ensures that no runtime errors will be encountered when executing the program, 
e.g. division by zero and numeric overflow/underflow. 
Verification at the next level, ``gold level,''  consists of verifying key user specifications, such as 
type invariants and subprogram preconditions and postconditions; 
however, at this level, the specifications only partially specify the desired behavior of the code.  
Verification at the next level, ``platinum level,'' consists verifying a complete set of specifications.
At the moment, the SPARK code generated by Salty proves at gold level.  
SPARK is able to prove the system initial and system transition property portions of the original GR(1) specification. 
Proof of system transition properties in turn requires proving a type invariant on the underlying Moore machine representation of the controller; 
this type invariant encodes that each state is reached by a specific set of input values and produces a specific set of output values. 
The platinum level would be reached if the liveness properties were expressed and proved, 
which we leave for future work.

In what follows, in \sectref{sec:implementation} we describe the structure of synthesized controller implementations in SPARK 
by walking through a simple example involving a traffic light. 
To evaluate whether verification of synthesized SPARK controllers is feasible and has utility, 
we pulled in a large body of examples from various sources. 
In \sectref{sec:vipExample}, we take a detailed look at one of these examples, a controller that coordinates the actions of 
a team of unmanned air vehicles (UAVs) performing an escort mission, 
and we describe how SPARK revealed an error in the controller's specification. 
In \sectref{sec:caseStudies}, we give an overview of results for the rest of the examples, with a focus on scalability. 
In \sectref{sec:lessonsLearned}, we discuss lessons learned, including how to structure synthesized controllers for proof.  
% examples in which SPARK revealed unexpected controller behaviors, and caveats related to GR(1) specifications. 
We end with concluding remarks in \sectref{sec:conclusions}, including a discussion of future work.

@@

We have extended the open source tool Salty\footnote{https://github.com/GaloisInc/salty/} to produce software implementations of controllers synthesized from GR(1) specifications in SPARK. 
Given a GR(1) specification in Salty format, Salty does some preprocessing to sanity check, optimize, and translate the specification to Slugs format. 
Salty then calls Slugs to synthesize a controller design from the processed specification. 
Slugs provides the option of returning the controller design as a Moore machine expressed in text format, 
which is what Salty uses to create an executable software implementation of the controller. 
This Moore machine encodes all the states of the controller. %, including a single initial state. 
It also encodes the transition relation between states, i.e. the controller's next state given the current state 
and the next set of input values. 
Note that this transition relation is deterministic, i.e. there is at most one possible next state for each state and set of input values. 
Every state then encodes a set of output values to be produced each time the controller transitions to that state. 
As in all Moore machines, each state produces exactly one set of output values. 
Controllers synthesized from GR(1) specifications have the additional property that for each state, 
the set of input values that brings the controller into that state is the same, regardless of what the previous state was.
% there is exactly one set of input values that brings the controller into that state.

If SPARK is chosen as the target progamming language, Salty translates this textual Moore machine representation 
of the controller to an implementation in SPARK. 
To give a brief overview, the Moore machine representing the controller is encoded as a record (analogous to a struct in C) 
that stores the controller's current state and a copy of the current input and output values, 
i.e. the inputs and outputs associated with the most recent transition. 
There is then a ``move'' procedure that implements the transition relation. 
This procedure changes the controller's state based on the next set of input values, produces the next set of output values, 
and updates its internal copy of current input and output values based on these most recent input and output values. 
Salty also synthesizes all of the annotations necessary to encode the specifications of interest 
and verify that the controller satisfies those specifications. 
In terms of specifications, we seek to prove that the controller satisfies the system initial properties $\varphi^s_i$ 
and system transition properties $\varphi^s_t$. 
Recall that GR(1) specifications take the form 
$(\varphi^e_i \land \varphi^e_t \land \varphi^e_l)  \limplies (\varphi^s_i \land \varphi^s_t \land \varphi^s_l)$. 
Note however that for controllers synthesized from GR(1) specifications,
system initial and transition properties should hold regardless of whether environment liveness properties hold. 
We are therefore interested in showing that 
$(\varphi^e_i \land \varphi^e_t)  \limplies (\varphi^s_i \land \varphi^s_t)$. 
For the SPARK implementation, Salty therefore generates functions corresponding to $\varphi^e_i$ and $\varphi^e_t$,
which are used as a precondition for the move procedure. 
It also generates functions corresponding to $\varphi^s_i$ and $\varphi^s_t$, 
which are used as a postcondition for the move procedure. 
If SPARK can verify that the move function satisfies the postcondition given the precondition, 
this verifies that $(\varphi^e_i \land \varphi^e_t)  \limplies (\varphi^s_i \land \varphi^s_t)$. 
In terms of additional annotations needed to prove this property in SPARK, Salty generates 
``state to input mapping'' and ``state to output mapping'' functions that encode which input values 
and output values are associated with each state; these are used to define an invariant for the controller.  
This is essentially all this is required to prove system initial and system transition properties of synthesized controllers in SPARK.

To explain synthesized SPARK controllers and annotations in more detail, we consider a simple example involving a traffic light controller. 
Let the controller's single input be $tick$. 
Color changes occur in every state in which $tick$ is true.
The traffic light's color cycles infinitely over the sequence red $\rightarrow$ green $\rightarrow$ yellow $\rightarrow$ red $\rightarrow$ $\ldots$.
Let the output variables then be $red$, $yellow$, and $green$.
The environment specifications are $\varphi^e_i = \top$, $\varphi^e_t = \top$, $\varphi^e_l = \always \eventually tick$, i.e.  
$tick$ can be true or false in the initial state and has no constraints on how it transitions from state to state, but it must be true infinitely often.
The system initial and liveness specifications are $\varphi^s_i = red \land \lnot yellow \land \lnot green$ and $\varphi^s_l =  \always \eventually green$, i.e.
the light starts as red but should infinitely often be green. The system transition specification is
%{\footnotesize
%\begin{align}
%\always & (\lnext ( (red \land \lnot yellow \land \lnot green) \lor (\lnot red \land yellow \land \lnot green) \lor \mbox{} \label{eq:mutex1} \\
%	& \quad \quad \quad (\lnot red \land \lnot yellow \land green)) \land \mbox{}  \label{eq:mutex2} \\
%	& \;\; (red \land \lnext tick \limplies \lnext green) \land (green \land \lnext tick \limplies \lnext yellow) \land \mbox{} \label{eq:change1} \\
%	& \quad \quad \quad (yellow \land \lnext tick \limplies \lnext red) \land \mbox{}  \label{eq:change2} \\
%	& \;\; (red \land \lnext \lnot tick \limplies red) \land (green \land \lnext \lnot tick \limplies green) \land \mbox{} \label{eq:same1} \\
%	& \quad \quad \quad (yellow \land \lnext \lnot tick \limplies yellow) \label{eq:same2})
%\end{align}
%}%
%That is, \eqnref{eq:mutex1} - \eqnref{eq:mutex2} express that the light should only be one color at a time; 
%\eqnref{eq:change1} - \eqnref{eq:change2} that the color should change from red to green to yellow and back to red whenever $tick$ is true; 
%and the color should remain the same whenever $tick$ is false.

@@

In SPARK, subunits consist of a specification and a body.
% For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
The package specification for this traffic light example is shown in \figref{fig:trafficLightSpec}.
Type \lstinline{Controller} encodes a Moore machine representing the synthesized controller. 
In the public part of the specification, \lstinline{Controller} (line 2) is declared as a private type so that the user cannot arbitrarily manipulate its state.  
% (possibly violating the controller's GR(1) specification). 
In the private part of the specification, \lstinline{Controller} (lines 33-37) is a record that stores the controller's internal state number, 
current input value(s), and current output value(s). 
Its initial state is always the last possible state number. 
In this example, each input and output is of type \lstinline{Boolean} (Salty also allows for enumerations and integers, to be discussed later). 
When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{Environment} or \lstinline{System} (lines 4-6), respectively. 
Here, there is only one input, so it is not wrapped in a record. 
Since \lstinline{Controller} is a Moore machine, for each internal state, 
there is exactly one set of output values produced in that state. 
Furthermore, recall that controllers synthesized from GR(1) specifications have the additional property that 
there is exactly one set of input values that brought the controller into that state. 
Therefore, \lstinline{Controller} includes a type invariant (lines 36-37) that captures this property, 
where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} state whether the controller has the corresponding state number and input or output value(s), respectively. 
% They use lookup tables, \lstinline{State_To_Input_Equivalence} and \lstinline{State_To_Output_Equivalence}, defined in the package body, to do so.
Both of these functions are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
unless it is specified during compilation that they should be.
Executing ghost code is often used for debugging, e.g. to check whether an unproved property is actually true. 
In this case, removing the ghost code can save significant memory, 
since \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
internally encode large lookup tables that have to store information on every state of the controller, 
which is often thousands of states or even millions of states for the largest example in our database.
The logic for these functions is given in the body (not shown here), but for instance when \lstinline{C.State = 1}, 
\lstinline{State_To_Input_Mapping(C)} returns \lstinline{True} if and only if \lstinline{C.tick = False}, and 
\lstinline{State_To_Output_Mapping(C)} returns \lstinline{True} if and only if 
\sloppy{\lstinline{C.S = System'(red => True, yellow => False, green => False)}}, 
since \lstinline{tick = False} brings the system into state 1 and $red$ is true and $yellow$ and $green$ are false in state 1.

>>>>

% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Institute 1 \\
\email{author1@email.com} \and
Institute 2 \\
\email{author2@email.com}  \and
Institute 3 \\
 \email{author3@email.com}}
%
%\institute{Air Force Research Laboratory, WPAFB, OH 45433, USA \\
%\email{laura.humphrey@us.af.mil} \and
%LinQuest Corp., Beavercreek, OH 45431, USA \\
%\email{james.hamil.ctr@us.af.mil}  \and
%AdaCore, F-75009 Paris, France \\
% \email{huguet@adacore.com}}
%
\maketitle              % typeset the header of the contribution
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ABSTRACT
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
%The abstract should briefly summarize the contents of the paper in 150--250 words.
Manually designing control logic for reactive systems is time-consuming and error-prone. 
An alternative is to automatically generate controllers using ``correct-by-construction'' synthesis approaches. 
Recently, there has been interest in synthesis from Generalized Reactivity(1) or GR(1) specifications, 
since the required computational complexity is relatively low, and several tools exist for synthesis from GR(1) specifications.
However, while these tools implement synthesis approaches that are theoretically ``correct-by-construction,'' 
errors in tool implementation can still lead to errors in synthesized controllers. 
We are therefore interested in ``end-to-end'' verification of synthesized controllers with respect to their original GR(1) specifications.
Toward this end, we have modified Salty -- a tool that produces executable software implementations of controllers 
from GR(1) specifications in a variety of programming languages -- to produce implementations in SPARK.
SPARK is both a programming language and associated set of verification tools, so it has the potential to enable the ``end-to-end'' verification we desire.
In this paper, we discuss our experience to date using SPARK to implement controllers and verify them against 
a subset of properties comprising GR(1) specifications, namely system initial and system transition properties. 
We also discuss lessons learned about how to best encode controllers synthesized from GR(1) specifications in SPARK for verification, 
examples in which verification found unexpected controller behaviors, and 
caveats related to the interpretation of GR(1) specifications.

@@

Reactive systems must be capable of correctly responding to various inputs, 
e.g. originating from human users or events in the system's operational environment. 
The process of manually designing the control logic for such systems is both time-consuming and error prone. 
An alternative is to use ``correct-by-construction'' synthesis approaches to automatically generate 
a system's control logic directly from specifications, which can 
reduce both the amount of time needed for design and the likelihood of errors 
\cite{alur2016compositional}, \cite{fainekos2009temporal}, \cite{guo2014cooperative}, \cite{kupermann2001synthesizing}. 
In general, \emph{synthesis} is the process of automatically generating a design from a specification. 
More specifically, \emph{reactive synthesis} approaches generate designs in the context of an uncontrolled environment, 
assumptions about which are encoded in the specification.
Reactive synthesis approaches tend to have high computational complexity,  
so there is particular interest in synthesis from Generalized Reactivity(1) or GR(1) specifications, 
the complexity of which is only polynomial in the size of the game graph encoded by the specification \cite{bloem2012}.
Synthesis from GR(1) specifications has been used to generate digital circuits \cite{ehlers2012symbolically} 
and controllers for teams of unmanned vehicles \cite{apker2016}, ground robots \cite{kress2007s}, 
software-defined networks \cite{wang2013automated}, and aircraft power distribution systems \cite{xu2012case}, to name a few.

A GR(1) specification $\varphi$ takes the form $\varphi = \varphi^e \limplies \varphi^s$, 
where $\varphi^e$ encodes assumptions about the \emph{environment} in which a system is to operate,
and $\varphi^s$ encodes guarantees the \emph{system} should make under those assumptions \cite{Ehlers2016}. 
More specifically, $\varphi$ takes the form 
$\varphi = (\varphi^e_i \land \varphi^e_t \land \varphi^e_l)  \limplies (\varphi^s_i \land \varphi^s_t \land \varphi^s_l)$,
where $\varphi^e_i$ and $\varphi^s_i$ are \emph{initial} properties, 
$\varphi^e_t$ and $\varphi^s_t$ are \emph{transition} or safety properties, and 
$\varphi^e_l$ and $\varphi^s_l$ are \emph{liveness} properties. 
For inputs in the set $\inputs$ controlled by the environment 
and outputs in the set $\outputs$ produced by the system, terms are defined as:

@@

Several tools for synthesis from GR(1) specifications are available. 
For example, RATSY %(Requirements Analysis Tool with Synthesis) 
\cite{bloem2010ratsy} has a focus on circuit design and synthesizes designs 
encoded in BLIF, %(Berkeley Logic Interchange Format),
Verilog, and HIF. %(HDL Intermediate Format). 
Similarly, Anzu \cite{jobstmann2007anzu} produces circuit designs in Verilog.
LTLMoP 
%(Linear Temporal Logic MissiOn Planning)  
\cite{finucane2010ltlmop} is focused on control of robots modeled as hybrid systems, 
and it synthesizes designs as hybrid controllers with handler modules to help connect controllers to simulated or real-world systems. 
TuLiP 
%(Temporal Logic Planning) 
\cite{TuLiP2011} has a similar focus and can synthesize controller implementations in Python. 
Slugs 
%(SmalL bUt Complete GROne Synthesizer) 
\cite{Ehlers2016} is architected to allow users to tailor synthesis algorithms,
 e.g. to optimize criteria such as quick response, cost-optimality, and error-resilience, and it produces mathematical representations of controller designs.
 %% -- TODO: put more details about input types that Salty allows? About what optimizations it does?
Salty \cite{elliott2019salty} provides a front-end to Slugs that makes specifications easier to write and debug and 
a back-end that turns controller designs into executable software implementations in a variety of programming languages, 
including Python, C++, Java, and now SPARK. 
Though all of these tools implement synthesis algorithms that are theoretically ``correct-by-construction,''
tool implementation errors could still result in errors in synthesized controllers.
We are therefore interested in ``end-to-end'' verification of synthesized controllers.
Toward this end, we have extended Salty to produce software implementations of synthesized controller designs in SPARK.

SPARK, which is based on the Ada programming language, is both a programming language 
with a specification language and an associated verification toolset \cite{hoang2015spark}.
% SPARK translates code to the Why language for processing by the Why3 theorem prover, 
% and Why3 uses other automated provers -- including SMT solvers Z3, CVC4, and Alt-Ergo -- to discharge verification conditions.
Though SPARK aims to perform fully automated verification, manual adaptation of the source code is in general necessary. 
For example, if one were to translate source code originally written in Ada to SPARK, 
one would have to remove unsupported features such as functions with side-effects, aliased variable names, exception handlers, etc.  
Formal verification of user-specified properties also requires manually writing specifications at the level of the source code in the 
form of contracts, e.g. preconditions and postconditions that summarize the assumptions and guarantees provided by individual subprograms.
There are different levels to which one can use SPARK to verify a program \cite{moy2019climbing}. 
Generally, the level of verification performed during development is incremental, going from lowest to highest. 
Verification at the lowest level, colloquially referred to as ``stone level,'' is achieved when code is accepted by SPARK, 
since the SPARK language has stricter legality rules than Ada. 
Verification at the next level, ``bronze level,'' is achieved when flow analysis returns with no error; 
flow analysis checks for errors such as reads of uninitialized data or violations of user-specified data flow contracts. 
Verification at the next level, ``silver level,'' ensures that no runtime errors will be encountered when executing the program, 
e.g. division by zero or numeric overflow/underflow. 
Verification at the next level, ``gold level,''  consists of verifying key user specifications, such as 
type invariants and subprogram preconditions and postconditions; 
however, at this level, the specifications only partially specify the desired behavior of the code.  
Verification at the next level, ``platinum level,'' consists verifying a complete set of specifications.
At the moment, the SPARK code generated by Salty proves at gold level.
In particular, SPARK is able to prove the system initial and system transition portions the original GR(1) specification. 
Proof of system transition properties in turn requires proving a type invariant on the underlying Moore machine representation of the controller; 
this type invariant encodes that each state is reached by a specific set of input values and produces a specific set of output values. 
Platinum level would be reached if the liveness properties were expressed and proved, 
which we leave for future work.

In what follows, in \sectref{sec:implementation} we describe the structure of synthesized controller implementations in SPARK 
by walking through a simple example involving a traffic light. 
To evaluate whether verification of synthesized SPARK controllers is feasible and has utility, 
we collected a corpus of examples from various sources. 
In \sectref{sec:vipExample}, we take a detailed look at one of these examples, a controller that coordinates the actions of 
a team of unmanned air vehicles (UAVs) performing an escort mission, 
and we describe how SPARK revealed an error in the controller's specification. 
In \sectref{sec:caseStudies}, we give an overview of results for the rest of the examples, with a focus on scalability. 
In \sectref{sec:lessonsLearned}, we discuss lessons learned, including how to best structure synthesized controllers for proof.  
% examples in which SPARK revealed unexpected controller behaviors, and caveats related to GR(1) specifications. 
We end with concluding remarks in \sectref{sec:conclusions}, including a discussion of future work.

@@

We have extended the open source tool Salty\footnote{https://github.com/GaloisInc/salty/} to produce software implementations of controllers synthesized from GR(1) specifications in SPARK. 
Given a GR(1) specification in Salty format, Salty does some preprocessing to sanity check, optimize, and translate the specification to Slugs format. 
Salty then calls Slugs to synthesize a controller design from the processed specification. 
Slugs provides the option of returning the controller design as a Moore machine expressed in text format, 
which is what Salty uses to create an executable software implementation of the controller. 
This Moore machine encodes all the states of the controller. %, including a single initial state. 
It also encodes the transition relation between states, i.e. the controller's next state given the current state 
and the next set of input values. 
% Note that this transition relation is deterministic, i.e. there is at most one possible next state for each state and set of input values. 
% Actually -- this is not true. There are examples where non-deterministic transitions are possible. In this case, Salty picks one arbitrarily.
Every state then encodes a set of output values to be produced each time the controller transitions to that state. 
As in all Moore machines, each state produces exactly one set of output values. 
Controllers synthesized from GR(1) specifications have the additional property that for each state, 
there is a unique set of input values that brings the controller into that state, regardless of what the previous state was.
% there is exactly one set of input values that brings the controller into that state.

If SPARK is chosen as the target programming language, Salty translates this textual Moore machine representation 
of the controller to an implementation in SPARK. 
To give a brief overview, the Moore machine representing the controller is encoded as a record (analogous to a struct in C) 
that stores the controller's current state and a copy of the current input and output values, 
i.e. the inputs and outputs associated with the most recent transition. 
There is then a ``move'' procedure that implements the transition relation. 
This procedure changes the controller's state based on the next set of input values, produces the next set of output values, 
and updates its internal copy of current input and output values based on these most recent input and output values. 
Salty also synthesizes all of the annotations necessary to encode the specifications of interest 
and verify that the controller satisfies those specifications. 
In terms of specifications, we seek to prove that the controller satisfies the system initial properties $\varphi^s_i$ 
and system transition properties $\varphi^s_t$. 
Recall that GR(1) specifications take the form 
$(\varphi^e_i \land \varphi^e_t \land \varphi^e_l)  \limplies (\varphi^s_i \land \varphi^s_t \land \varphi^s_l)$. 
Note however that for controllers synthesized from GR(1) specifications,
system initial and transition properties should hold regardless of whether environment liveness properties hold. 
We are therefore interested in showing that 
$(\varphi^e_i \land \varphi^e_t)  \limplies (\varphi^s_i \land \varphi^s_t)$. 
For the SPARK implementation, Salty therefore generates functions corresponding to $\varphi^e_i$ and $\varphi^e_t$,
which are used in the precondition for the move procedure. 
It also generates functions corresponding to $\varphi^s_i$ and $\varphi^s_t$, 
which are used in the postcondition for the move procedure. 
If SPARK can verify that the move function satisfies the postcondition given the precondition, 
this verifies that $(\varphi^e_i \land \varphi^e_t)  \limplies (\varphi^s_i \land \varphi^s_t)$. 
In terms of additional annotations needed to prove this property in SPARK, Salty generates 
``state to input mapping'' and ``state to output mapping'' functions that encode which input values 
and output values are associated with each state; these are used to define a type invariant on the controller.  
This is essentially all that is required to prove system initial and system transition properties of synthesized controllers in SPARK.

To understand synthesized SPARK controllers and annotations in more detail, consider a simple example involving a traffic light controller. 
Let the controller's single input be $tick$. 
Color changes occur in every state in which $tick$ is true.
The traffic light's color cycles infinitely over the sequence red $\rightarrow$ green $\rightarrow$ yellow $\rightarrow$ red $\rightarrow$ $\ldots$.
Let the output variables then be $red$, $yellow$, and $green$.
The environment specifications are $\varphi^e_i = \top$, $\varphi^e_t = \top$, $\varphi^e_l = \always \eventually tick$, i.e.  
$tick$ can be true or false in the initial state and has no constraints on how it transitions from state to state, but it must be true infinitely often.
The system initial and liveness specifications are $\varphi^s_i = red \land \lnot yellow \land \lnot green$ and $\varphi^s_l =  \always \eventually green$, i.e.
the light starts as red but should infinitely often be green. The system transition specification is
%{\footnotesize
%\begin{align}
%\always & (\lnext ( (red \land \lnot yellow \land \lnot green) \lor (\lnot red \land yellow \land \lnot green) \lor \mbox{} \label{eq:mutex1} \\
%	& \quad \quad \quad (\lnot red \land \lnot yellow \land green)) \land \mbox{}  \label{eq:mutex2} \\
%	& \;\; (red \land \lnext tick \limplies \lnext green) \land (green \land \lnext tick \limplies \lnext yellow) \land \mbox{} \label{eq:change1} \\
%	& \quad \quad \quad (yellow \land \lnext tick \limplies \lnext red) \land \mbox{}  \label{eq:change2} \\
%	& \;\; (red \land \lnext \lnot tick \limplies red) \land (green \land \lnext \lnot tick \limplies green) \land \mbox{} \label{eq:same1} \\
%	& \quad \quad \quad (yellow \land \lnext \lnot tick \limplies yellow) \label{eq:same2})
%\end{align}
%}%
%That is, \eqnref{eq:mutex1} - \eqnref{eq:mutex2} express that the light should only be one color at a time; 
%\eqnref{eq:change1} - \eqnref{eq:change2} that the color should change from red to green to yellow and back to red whenever $tick$ is true; 
%and the color should remain the same whenever $tick$ is false.

@@

In SPARK, subunits consist of a specification and a body.
% For each synthesized controller, Salty automatically generates both a specification and body for a package named according to the Salty specification. 
The package specification for this traffic light example is shown in \figref{fig:trafficLightSpec}.
Type \lstinline{Controller} encodes a Moore machine representing the synthesized controller. 
In the public part of the specification, \lstinline{Controller} (line 2) is declared as a private type so that the user cannot arbitrarily manipulate its state.  
% (possibly violating the controller's GR(1) specification). 
In the private part of the specification, \lstinline{Controller} (lines 33-37) is a record that stores the controller's internal state number, 
current input value(s), and current output value(s). 
The internal state number is always initialized to the last possible state number; 
this ``controller initialization'' state encodes the status of the controller before any inputs are received, and the only transition out of this state are those allowed by $\varphi_i^e$.
In this example, each input and output is of type \lstinline{Boolean} (Salty also allows for enumerations and integers, to be discussed later). 
When there are multiple inputs or outputs, they are wrapped in a record of type \lstinline{Environment} or \lstinline{System} (lines 4-6), respectively. 
Here, there is only one input, so it is not wrapped in a record. 
Since \lstinline{Controller} is a Moore machine, for each internal state, 
there is exactly one set of output values produced in that state. 
Furthermore, recall that controllers synthesized from GR(1) specifications have the additional property that 
there is exactly one set of input values that brings the controller into each state. 
\lstinline{Controller} therefore includes a type invariant (lines 36-37) that captures this property, 
where functions \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} evaluate 
whether the controller's state corresponds to the expected input and output value(s), respectively. 
% They use lookup tables, \lstinline{State_To_Input_Equivalence} and \lstinline{State_To_Output_Equivalence}, defined in the package body, to do so.
Both of these functions are declared (lines 26-29) with aspect \lstinline{Ghost}, indicating that they are intended mainly for proof purposes, 
i.e. they will create verification conditions related to type \lstinline{Controller} but will not be executed in the actual program, 
unless it is specified during compilation that they should be.
Executing ghost code is often used for debugging, e.g. to check an unproved property through testing. 
In this case, removing the ghost code can save significant memory, 
since \lstinline{State_To_Input_Mapping} and \lstinline{State_To_Output_Mapping} 
internally encode large lookup tables that have to store information on every state of the controller, 
which is often thousands of states or even millions of states for the largest example in our database.
The logic for these functions is given in the body (not shown here), but for instance when \lstinline{C.State = 1}, 
\lstinline{State_To_Input_Mapping(C)} returns \lstinline{True} if and only if \lstinline{C.tick = False}, and 
\lstinline{State_To_Output_Mapping(C)} returns \lstinline{True} if and only if 
\sloppy{\lstinline{C.S = System'(red => True, yellow => False, green => False)}}, 
since \lstinline{tick = False} brings the system into state 1, and $red$ is true and $yellow$ and $green$ are false in state 1.
